Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2016-07-04 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
Included CF comments through HRR
also bib typos fixed along the way

The need for fire investigation and value to society, tools used , example.

\subsection{The need and purpose for Fire Investigation}
Unwanted fires result in loss of life, property and can create an adverse economic impact on a community.  Recent findings from an archaeological site in South Africa, indicate that mankind has productively used fire for approximately 1 million years~\cite{Berna:2012}.  This is 300 thousand years earlier than previously thought.  Despite mankind's extensive history with fire, according to the World Health Organization, an estimated 265,000 people die each year as a result of burn injuries~\cite{WHO:2014}.  Based on an estimated global population of 7.3 billion people, the death rate due to burns from all sources would be 3.6 per 100,000 people.  The World Fire Statistics Center collects data from European, North American and a few Asia-Oceania countries.  Within this group of countries the fire deaths per 100,000 population range from 0.02 (Singapore) to 2.03 (Finland)~\cite{Climate:2014}. There are a significant number of factors that determine the variability of fire death rates between nations, socio-economic conditions being a leading factor~\cite{WHO:2014}.

@@

In some cases, the fire investigation determines that the fire was intentional.  An intentional fire as defined in the National Fire Incident Reporting System (NFIRS)  are those fires that are deliberately set and include fires that result from deliberate misuse of a heat source, fires of an incendiary nature (arson), as well as controlled burn fires, such as crop clearing, that required fire service intervention~\cite{Campbell:2014}.  An incendiary fire, as defined by the National Fire Protection Association (NFPA) Guide for Fire \& Explosion Investigations, commonly referred to as NFPA~921, is ``a fire that is deliberately set with the intent to cause a fire to occur in an area where the fire should not be''~\cite{NFPA_921}.

According to the NFPA, there were approximately 282,600 intentionally set fires each year during the period of 2007 through 2011 in the United States. The average annual loss totals for that period were approximately 420 civilian deaths and 1,360 civilian injuries. More than \$1.3 billion dollars (U.S.) was lost per year due to direct property damage. Fires involving structures account for more than 84~\% of the civilian deaths, injuries and property loss from all intentional fires.  Approximately two-thirds of the structure fires occurred in residential occupancies.  95~\% of the civilian deaths and 86~\% of the civilian injuries also occurred in the intentional fires in residential occupancies~\cite{Campbell:2014}.

A subset of the intentional fires are considered arson.  Arson as defined by the U.S. Federal Bureau of Investigation (FBI) Uniform Crime Reporting (UCR) Program defines arson as any willful or malicious burning or attempting to burn, with or without intent to defraud, a dwelling house, public building, motor vehicle or aircraft, personal property of another, etc.~\cite{Crime:2010}.

The national rate of arson fire cases cleared by arrest or exceptional means is less than 20~\%.  Of those cases approximately 46~\% of the suspects were prosecuted~\cite{Campbell:2014}. Arson is a challenging crime to solve because typically there are no witnesses present at the time of the crime.

The values cited above are believed to be the best available in the United States and perhaps one of the better fire incident databases in the world.  However an investigative report by Icove and Hargrove, indicates that arson fires are underreported to the National Fire Incident Reporting System (NFIRS).  The investigation provided examples from several major cities, such as New York City, where in 2011, 11 arson fires in buildings were reported to NFIRS, while the Fire Department of the City of New York's Bureau of Fire Investigation had recorded 1,347 arson fires in buildings that year.  Similar discrepancies were found in eight other large U.S. cities.  On average, three-fourths of the arsons uncovered by city fire investigators went unreported to NFIRS~\cite{Icove_2014}.

Fires are investigated in order to determine the ``origin and cause'' of the fire and in cases of arson there is an effort to determine who was responsible for setting the fire.  This investigation process must follow the scientific method as documented in NFPA 921~\cite{NFPA_921}.  There are numerous textbooks on the subject of conducting a fire investigation~\cite{Almirall:2004,Fire_Investigation,DeHaan:2012,Icove:2013,Lentini:2006,Noon:1995}. The NFPA guide and the texts provide information to a fire investigator collecting data from the scene, analyzing the data, and developing a hypothesis about the fire.

Patterns produced by the fire are in many cases a significant portion of the data collected and are analyzed at the scene to determine the area of origin.  One of the basic methods of documenting the fire scene is to photograph fire patterns.  As noted by Icove, DeHaan, and Haynes  ``'the ability to document and interpret fire patterns accurately is essential to investigators reconstructing fire scenes…''~\cite{Icove:2013}.  A fire pattern is defined in NFPA~921 as, ``the visible or measurable physical change, or identifiable shapes, formed by a fire effect or group of fire effects''~\cite{NFPA_921}.

@@

Previous fire pattern research by the National Institute of Justice (NIJ), the National Institute of Standards and Technology (NIST), and the United States Fire Administration (USFA) has shown that fire patterns provide data useful for the determination of fire origin.  The reports noted the impact of ventilation on the development of the burn patterns~\cite{Shanley:1997,Putorti:1997}. A large number of other factors affect the formation of these patterns: burn time, heat release rate of the fire source, fire exposure, target fuel composition, adjacent fuel(s) and compartmentalization, to name a few. Given the limited number of experiments in the literature and the large number of variables, it has been difficult to fully understand a cause and effect relationship between the fire scenarios and the resulting patterns.

Fire models are another tool that may be used by a fire investigation team to gain insight into the growth and development of a fire and to test the investigator’s hypothesis on origin and cause~\cite{Sutula}.  Many theoretical and empirical correlations and computational fire models have been developed for use in fire protection design, fire hazard analysis, and fire investigation.  A collection of the widely used correlations can be found in the U.S. Nuclear Regulatory Commission’s (NRC) Fire Dynamics Tools (FDTs). These spread sheets incorporate references to the source documents and experimental work upon which the correlations are based. In addition, NRC tasked NIST with conducting a verification and validation study of the correlations. Overholt developed the NIST verification and validation study~\cite{Overholt:2014}. Not all of the correlations in the FDTs were included in the study because an insufficient amount of reliable data is available to conduct validation.  Two correlations that could not be validated were for flame height and time to ignition.

@@

The fire investigation community recognized a need to improve the science and practice of fire investigation.  One of the manifestations of the efforts to improve the practice was the development of NFPA~921, {\em Guide for Fire and Explosion Investigations}~\cite{NFPA_921}.  The first edition of NFPA~921 was issued in 1992.  As provided by the document scope, it ``is designed to assist individuals who are charged with the responsibility of investigating and analyzing fire and explosion incidents and rendering opinions as to the origin, cause, responsibility, or prevention of such incidents, and the damage and injuries which arise from such incidents.'' The document’s purpose includes the goal to be ``a model for the advancement and practice of fire and explosion investigation, fire science, and methodology.'' However, NFPA~921 can only incorporate the fire science and research results that have been produced.  Although NFPA~921 is a ``guide'' as opposed to a ``standard,'' it is considered the standard of care for the fire investigation community in countries that recognize and use NFPA standards and guides.

@@

Given that the majority of intentional fires in structures occurred in residential occupancies, this study will focus on fire patterns on painted gypsum wallboard, which is 2.4~m in height.  Given the upper bound of 2.4~m, the experiments will use small fires with mean flame heights of approximately 1.2~m.  The resulting patterns on the gypsum board wall would be representative of a pre-flashover fire pattern.

@@

@@

A series of experiments were conducted to characterize a limited set of  residential scale, pre-flashover or incipient fires.  A fixed fire source footprint, 30.5~cm square, was chosen for this study and a range of common fuels was selected.  The fires were positioned under an oxygen consumption calorimetry hood to determine the heat release rate.  Heat release rate, flame/plume gas temperatures, and total heat flux from the flames were measured and flame movement and height were recorded with photographs and videos.  Given the focus on residential fires, small source fires were selected so that the burn pattern would be limited to less than 2.4~m in height.  Replicate experiments were conducted with each of the fuels, in order to examine the repeatability of the fires and to quantify the variability in the flames.

@@

Three different fuels were used in this study.  Experiments were conducted with a natural gas fueled burner, a liquid fueled pan fire, and a solid fuel. Natural gas was chosen as a fuel because it is used for calibrating the oxygen depletion calorimeters in NIST’s Large Fire Laboratory.  Therefore the heat content of the gas was monitored and mass flow controlled to provide a heat release rate based on fuel mass flow for comparison with the heat release rate values measured with the calorimeter.  The gas burner arrangement provides the most idealized fire, in that it can be ignited and brought up to a near steady state heat release rate within a few seconds and then be shut down just as quickly.  This provided an approximate `square wave' exposure to the gypsum wallboard.  This is important for determining the potential amount of energy transferred to the exposed surface.

@@

The source fire characterization experiments were conducted in the NIST Large Fire Research Laboratory's 3~m by 3~m hood oxygen consumption calorimeter. This apparatus follows the methodology of ASTM~E 2067-08, {\em Standard Practice for Full-Scale Oxygen Consumption Calorimetry Fire Tests} ~\cite{ASTM_E2067}. The fires were centered under the hood of the calorimeter and positioned in an open steel frame which provided support for the thermocouples and heat flux sensors. A load cell was positioned as a support for the fuels.  All of the data was recorded at intervals of 1~s on a computer based data acquisition system.

@@

The NIST Large Fire Research Laboratory's 3~m by 3~m oxygen depletion calorimeter has an estimated peak heat release rate capacity of approximately 700~kW.  Details on the operation and uncertainty in measurements associated with NIST’s 6~m by 6~m oxygen consumption calorimeter were documented by Bryant et al.~\cite{Bryant:2004}. Many of the same instruments are shared between the two calorimeters and calibration burns were conducted during each experimental series.  As a result, the expanded uncertainty of this device has been estimated as 11~\% on the measured heat release rate.

@@

Gas temperatures were measured with exposed-bead, Chromel-Alumel (type K) thermocouples, with a 1.0~mm nominal diameter.  Starting from the exposed bead, the thermocouple wire was sheathed in a 3.2~mm diameter inconel shield, 0.76~m in length.  The standard uncertainty in temperature of the thermocouple wire itself is 2.2~$^\circ$C at 277~$^\circ$C and increases to 9.5~$^\circ$C at 871~$^\circ$C as determined by the wire manufacturer~\cite{Omega}.  The variation of the temperature in the environment surrounding the thermocouple is known to be much greater than that of the wire uncertainty~\cite{Blevins:1999,Pitts:2001}. Small diameter thermocouples were used to limit the impact of radiative heating and cooling.  The estimated combined expanded uncertainty for temperature in these experiments is 15~\%.

@@

The heat flux was measured with water cooled, Schmidt-Boelter total heat flux gauges. The sensing surface of the heat flux gauges were located 0.95~m horizontally from the center of the burner/fuel and 0.50~m above the top of the burner/pan lip/fuel surface.  The manufacturer reports a 3~\% calibration expanded uncertainty for these devices~\cite{Medtherm}. Results from an international study on total heat flux gauge calibration and response demonstrated that the uncertainty of a Schmidt-Boelter gauge is typically 8~\%~\cite{Pitts:2006}.  The gauges were mounted to uprights of the steel instrumentation support frame on opposite sides of the fire.

@@

The fires were centered under the hood of the calorimeter on a non-combustible platform measuring 0.91~m x 1.2~m x 12~mm thick supported by the load cell.  The load cell had a resolution of 1.0~g~\cite{Mettler}. The expanded uncertainty is estimated to be 5~\%.  The load cell was checked between experiments with calibrated weights during the zero and span checks of the calorimeter.

@@

A 35~mm single lens reflex camera was used to record digital still photographs of the fires.  The camera had an 8 megapixel resolution.  A mini DV camcorder with 0.7 megapixel resolution was used to capture the video.  A firefighting type thermal imager with 0.08 megapixel resolution was used to capture infrared images in the 8 to 14 micron range.  Each camera was positioned on a tripod with a distance of 4.5~m between the centerline of burner to camera lens 4.5~m.  The height of each camera was adjusted so that the center of camera lens was positioned 0.5~m above top of the burner.

@@

On each fuel’s averaged heat release rate curve, the filled area surrounding the average represents the 95~\% confidence level (2$\sigma$), based on Type A evaluation of the uncertainty is provided~\cite{Taylor:1994}.  A Type A evaluation of the uncertainty is any statistical analysis of the data from a series of observations.  In this case the evaluation assumed a normal distribution of the data such that doubling the calculated standard deviation (2$\sigma$) would yield a confidence level of approximately 95~\%.

The natural gas burner provided the most repeatable results, with an average steady state heat release rate of 80~kW~$\pm$~9~kW with an average total heat release of 23.3~MJ~$\pm$~1.6~MJ or 11~\% over a 100~s period. It should be noted that 11~\% is also the (2$\sigma$) uncertainty estimate of the calorimeter.  The gasoline, based on the average over a 30 s interval bounding the peak heat release rate of each of the 18 experiments, provided an average peak heat release rate of 80~kW~$\pm$~24~kW or 20~\%.  The total heat release of the gasoline over the 300~s period after ignition was 13.6~MJ~$\pm$~0.6~MJ.  The polyurethane foam demonstrated the largest variability of the three fuels in terms of peak heat release rate and time to peak heat release.  Averaging over a 15~s interval bounding the peak heat release rate from each of the ten experiments yielded an average peak heat release rate of 45~kW~$\pm$~21~kW or 34~\%.  The total heat release over the 300~s period after ignition was 4.1~MJ~$\pm$~0.2~MJ

>>>>

This section reviews the fire problem to demonstrate the need and value to society that fire investigation can provide. Following the review, a brief presentation on the analysis  of fire patterns, the use of analytical tools, and the current challenges to the field of fire forensics is given as they are the reason for this study.  In closing this section, an example of a fire incident is presented. In the closing section we will return to the example again.

\subsection{The Need for Fire Investigation}
Unwanted fires result in loss of life, property and can create an adverse economic impact on a community.  Recent findings from an archaeological site in South Africa, indicate that mankind has productively used fire for approximately 1 million years~\cite{Berna:2012}.  This is 300 thousand years earlier than previously thought.  Despite mankind's extensive history with fire, according to the World Health Organization, an estimated 265,000 people die each year as a result of burn injuries~\cite{WHO:2014}.  Based on an estimated global population of 7.3 billion people, the death rate due to burns from all sources would be 3.6 per 100,000 people.  The World Fire Statistics Center collects data from European, North American and a few Asia-Oceania countries.  Within this group of countries the fire deaths per 100,000 population range from 0.02 (Singapore) to 2.03 (Finland)~\cite{Climate:2014}. There are a significant number of factors that determine the variability of fire death rates between nations, socio-economic conditions being a leading factor~\cite{WHO:2014}.

@@

In some cases, the fire investigation determines that the fire was intentional.  An intentional fire as defined in the National Fire Incident Reporting System (NFIRS)  are those fires that are deliberately set and include fires that result from deliberate misuse of a heat source, fires of an incendiary nature (arson), as well as controlled burn fires, such as crop clearing, that required fire service intervention~\cite{Campbell:2014}.  An incendiary fire, as defined by the National Fire Protection Association (NFPA) Guide for Fire \& Explosion Investigations, commonly referred to as NFPA~921, is ``a fire that is deliberately set with the intent to cause a fire to occur in an area where the fire should not be''~\cite{NFPA:921}.

According to the NFPA, there were approximately 282,600 intentionally set fires each year during the period of 2007 through 2011 in the United States. The average annual loss totals for that period were approximately 420 civilian deaths and 1,360 civilian injuries. More than \$1.3 billion dollars (U.S.) was lost per year due to direct property damage. Structure fires represent less than 20~\% of intentionally set fires.  However, fires involving structures account for more than 84~\% of the civilian deaths, injuries and property loss from all intentional fires.  Approximately two-thirds of the intentional structure fires occurred in occupied and operating residential occupancies.  As a result, 95~\% of the civilian deaths and 86~\% of the civilian injuries which are attributed to intentional fires occurred in residential occupancies~\cite{Campbell:2014}.

A subset of the intentional fires are considered arson.  The U.S. Federal Bureau of Investigation (FBI) Uniform Crime Reporting (UCR) Program defines arson as any willful or malicious burning or attempting to burn, with or without intent to defraud, a dwelling house, public building, motor vehicle or aircraft, personal property of another, etc.~\cite{Crime:2010}.  The national rate of arson fire cases cleared by arrest or exceptional means is less than 20~\%.  Of those cases approximately 46~\% of the suspects were prosecuted~\cite{Campbell:2014}.  Hence less than 10~\% of the arson cases resulted in successful prosecution.

The values cited above are believed to be the best available in the United States and perhaps one of the better fire incident databases in the world.  However an investigative report by Icove and Hargrove, indicates that arson fires are underreported to the U.S. National Fire Incident Reporting System (NFIRS).  The investigation provided examples from several major cities, such as New York City, where in 2011, 11 arson fires in buildings were reported to NFIRS, while the Fire Department of the City of New York's Bureau of Fire Investigation had recorded 1,347 arson fires in buildings that year.  Similar discrepancies were found in eight other large U.S. cities.  On average, three-fourths of the arsons uncovered by city fire investigators went unreported to NFIRS~\cite{Icove_2014}.  In other words, the number of arson related incidents is likely to be under represented in the fire incident data.

\subsection{The Use of Fire Patterns in Fire Investigation}
Fires are investigated in order to determine the ``origin and cause'' of the fire and in cases of arson there is an effort to determine who was responsible for setting the fire.  This investigation process must follow the scientific method as documented in NFPA 921~\cite{NFPA:921}.  There are numerous textbooks on the subject of conducting a fire investigation~\cite{Almirall:2004,Fire_Investigation,DeHaan:2012,Icove:2013,Lentini:2006,Noon:1995}. The NFPA guide and the texts provide information how to collect data from the scene, analyze the data, and develop a hypothesis about the fire.

Patterns produced by the fire are in many cases a significant portion of the data collected and are analyzed at the scene to determine the area of origin.  One of the basic methods of documenting the fire scene is to photograph fire patterns.  As noted by Icove, DeHaan, and Haynes  ``'the ability to document and interpret fire patterns accurately is essential to investigators reconstructing fire scenes…''~\cite{Icove:2013}.  A fire pattern is defined in NFPA~921 as, ``the visible or measurable physical change, or identifiable shapes, formed by a fire effect or group of fire effects''~\cite{NFPA:921}.

@@

Previous fire pattern research by the National Institute of Justice (NIJ), the National Institute of Standards and Technology (NIST), and the United States Fire Administration (USFA) has shown that fire patterns provide data useful for the determination of fire origin.  The reports noted the impact of ventilation on the development of the burn patterns~\cite{Shanley:1997,Putorti:1997}. A large number of other factors affect the formation of these patterns: burn time, heat release rate of the fire source, fire exposure, target fuel composition, adjacent fuel(s) and compartmentalization, to name a few. Given the limited number of experiments in the literature and the large number of variables, it has been difficult to develop a cause and effect relationship between the fire scenarios and the resulting patterns, based solely on the research.

Fire models are another tool that may be used by a fire investigation team to gain insight into the growth and development of a fire and to test the investigator’s hypothesis on origin and cause~\cite{Sutula:2001}.  Many theoretical and empirical correlations and computational fire models have been developed for use in fire protection design, fire hazard analysis, and fire investigation.  A collection of the widely used correlations can be found in the U.S. Nuclear Regulatory Commission’s (NRC) Fire Dynamics Tools (FDTs). These spread sheets incorporate references to the source documents and experimental work upon which the correlations are based. In addition, NRC tasked NIST with conducting a verification and validation study of the correlations. Overholt developed the NIST verification and validation study~\cite{Overholt:2014}. Not all of the correlations in the FDTs were included in the study because an insufficient amount of reliable data is available to conduct validation.  The correlations for flame height and time to ignition were not included in the validation study.

@@

The fire investigation community recognized a need to improve the science and practice of fire investigation.  One of the manifestations of the efforts to improve the practice was the development of NFPA~921, {\em Guide for Fire and Explosion Investigations}~\cite{NFPA:921}.  The first edition of NFPA~921 was issued in 1992.  As provided by the document scope, it ``is designed to assist individuals who are charged with the responsibility of investigating and analyzing fire and explosion incidents and rendering opinions as to the origin, cause, responsibility, or prevention of such incidents, and the damage and injuries which arise from such incidents.'' The document’s purpose includes the goal to be ``a model for the advancement and practice of fire and explosion investigation, fire science, and methodology.'' However, NFPA~921 can only incorporate the fire science and research results that have been produced.  Although NFPA~921 is a ``guide'' as opposed to a ``standard,'' it is considered the standard of care for the fire investigation community in countries that recognize and use NFPA standards and guides.

@@

Given that the majority of intentional fires in structures occurred in residential occupancies, this study will focus on fire patterns on painted gypsum wallboard which is a common interior finish for residential occupanies.  A common height for a residential occupanices is 2.4~m in height.  With a wall height of 2.4~m, these experiments will use small fires with mean flame heights of approximately 1.2~m.  The resulting patterns on the gypsum board wall would be representative of a pre-flashover fire pattern.

@@

The first set of fire pattern experiments on painted gypsum wallboard, without potentially intrusive instrumentation, was conducted to examine the impact of the construction on the wall. Three different types of wall construction were examined to determine if the amount of insulation behind the exposed gypsum wallboard changed the fire pattern by a significant amount.  The results would then be used to plan the wall construction of the remaining full- scale experiments.

@@

A series of experiments were conducted to characterize a limited set of  residential scale, pre-flashover or early growth stage fires.  A fixed fire source footprint, 30.5~cm square, was chosen for this study and a range of common fuels that could be found in and around a residence were selected.  The fires were positioned under an oxygen consumption calorimetry hood to determine the heat release rate.  Heat release rate, flame/plume gas temperatures, and total heat flux from the flames were measured and flame movement and height were recorded with photographs and videos.  Given the focus on residential fires, small source fires were selected so that the burn pattern would be limited to less than 2.4~m in height.  Replicate experiments were conducted with each of the fuels, in order to examine the repeatability of the fires and to quantify the variability in the flames.

@@

Three different fuels were used in this study.  Experiments were conducted with a natural gas fueled burner, a liquid fueled pan fire, and a solid fuel. Natural gas was chosen as a fuel for two reasons; it is a common fuel used for heating homes and it is used for calibrating the oxygen depletion calorimeters in NIST’s Large Fire Laboratory.  Therefore the heat content of the gas was monitored and mass flow controlled to provide a heat release rate based on fuel mass flow for comparison with the heat release rate values measured with the calorimeter.  The gas burner arrangement provides the most idealized fire, in that it can be ignited and brought up to a near steady state heat release rate within a few seconds and then be shut down just as quickly.  This provided an approximate `square wave' exposure to the gypsum wallboard.  This is important for determining the potential amount of energy transferred to the exposed surface.

@@

The source fire characterization experiments were conducted in the NIST Large Fire Research Laboratory's 3~m by 3~m hood oxygen consumption calorimeter. This apparatus follows the methodology of ASTM~E 2067-08, {\em Standard Practice for Full-Scale Oxygen Consumption Calorimetry Fire Tests} ~\cite{ASTM_E2067}. The fires were centered under the hood of the calorimeter and positioned in an open steel frame which provided support for the thermocouples and heat flux sensors. A load cell was positioned on the floor as a support for the fuels.  All of the data was recorded at intervals of 1~s on a computer based data acquisition system.

@@

The NIST Large Fire Research Laboratory's 3~m by 3~m oxygen depletion calorimeter had an estimated peak heat release rate capacity of approximately 700~kW.  Details on the operation and uncertainty in measurements associated with NIST’s 6~m by 6~m oxygen consumption calorimeter were documented by Bryant et al.~\cite{Bryant:2004}. Many of the same instruments were shared between the two calorimeters and calibration burns were conducted during each experimental series.  As a result, the expanded uncertainty of this device has been estimated as 11~\% on the measured heat release rate.

@@

Gas temperatures were measured with exposed-bead, Chromel-Alumel (type K) thermocouples, with a 1.0~mm nominal diameter.  Starting from the exposed bead, the thermocouple wire was sheathed in a 3.2~mm diameter inconel shield, 0.76~m in length. Thermocouple is ungrounded. The standard uncertainty in temperature of the thermocouple wire itself is 2.2~$^\circ$C at 277~$^\circ$C and increases to 9.5~$^\circ$C at 871~$^\circ$C as determined by the wire manufacturer~\cite{Omega}.  The variation of the temperature in the environment surrounding the thermocouple was known to be much greater than that of the wire uncertainty~\cite{Blevins:1999,Pitts:2001}. Small diameter thermocouples were used to limit the impact of radiative heating and cooling.  The estimated combined expanded uncertainty for temperature in these experiments was 15~\%.

@@

The heat flux was measured with water cooled, Schmidt-Boelter total heat flux gauges. The sensing surface of the heat flux gauges were located 0.95~m horizontally from the center of the burner/fuel and 0.50~m above the top of the burner/pan lip/fuel surface.  The manufacturer reports a 3~\% calibration expanded uncertainty for these devices~\cite{Medtherm}. Results from an international study on total heat flux gauge calibration and response demonstrated that the uncertainty of a Schmidt-Boelter gauge was typically 8~\%~\cite{Pitts:2006}.  The gauges were mounted to uprights of the steel instrumentation support frame on opposite sides of the fire.

@@

The fires were centered under the hood of the calorimeter on a non-combustible platform measuring 0.91~m x 1.2~m x 12~mm thick supported by the load cell.  The load cell had a resolution of 1.0~g~\cite{Mettler}. The expanded uncertainty was estimated to be 5~\%.  The load cell was checked between experiments with calibrated weights during the zero and span checks of the calorimeter.

@@

A 35~mm single lens reflex camera was used to record digital still photographs of the fires.  The camera had an 8 megapixel resolution.  A mini DV camcorder with 0.7 megapixel resolution was used to capture the video.  A firefighting type thermal imager with 0.08 megapixel resolution was used to capture infrared images in the 8 to 14 micron range.  Each camera was positioned on a tripod with a distance of 4.5~m between the centerline of burner to camera lens.  The height of each camera was adjusted so that the center of camera lens was positioned 0.5~m above top of the burner.

@@

On each fuel’s averaged heat release rate curve, the shaded area surrounding the average represents the 95~\% confidence level (2$\sigma$), based on Type A evaluation of the uncertainty is provided~\cite{Taylor:1994}.  A Type A evaluation of the uncertainty is any statistical analysis of the data from a series of observations.  In this case the evaluation assumed a normal distribution of the data such that doubling the calculated standard deviation (2$\sigma$) would yield a confidence level of approximately 95~\%.

The natural gas burner provided the most repeatable results, with an average steady state heat release rate of 80~kW~$\pm$~9~kW with an average total heat release of 23.3~MJ~$\pm$~1.6~MJ over a 100~s period. The uncertainty of the steady state heat release rate measurement is approximately~$\pm$~11~\% It should be noted that 11~\% is also the (2$\sigma$) uncertainty estimate of the calorimeter. The gas burner and on/off control valve allowed the heat release rate to be similar in form to a "square-wave".

The gasoline pan fire has was ignited and then burned until the fuel was consumed.  Therefore the heat release rate curves had more of a "bell shape"  as opposed to the flat steady- state peak heat release rate region of the natural gas burner.  Therefore the  gasoline average peak heat release rate was based on an average over a 30 s interval bounding the peak heat release rate of each of the 18 experiments.  This method provided an average peak heat release rate of 80~kW~$\pm$~24~kW or 30~\%.  The total heat release of the gasoline over the 300~s period after ignition was 13.6~MJ~$\pm$~0.6~MJ.

The block of polyurethane foam demonstrated the largest variability of the three fuels in terms of peak heat release rate and time to peak heat release. Similar to the gasoline that began with a fixed quantity of fuel which was ignitied and burned to completion, the polyurthane foam also generated "bell shaped" heat release rate curve.  Since the duration of the heat release rate peaks were less than those of the gasoline, the peak heat release rate averaging was performed over a 15~s interval bounding the peak heat release rate from each of the ten experiments.  This process yielded an average peak heat release rate of 45~kW~$\pm$~21~kW or 47~\%.  The total heat release over the 300~s period after ignition was 4.1~MJ~$\pm$~0.2~MJ.

The Figures~\ref{HRR}-\ref{HRR2} have a shaded area that shows the 95~\% confidence level.  The more narrow the shaded area, the better the repeatability of the fire.  This provides a quick means to assess the repeatability of a given source fire.  Using this measure the natural gas burner fires were the most repeatable, followed by the gasoline pan fires and then the polyurethane fires.
